---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**A1** Of the 17,083 in the dataframe, all but seven appear to be new registrants. Montgomery County, the state's most populated jurisdiction, leads with 4,847, more than 2,000 more than Baltimore City (which is the fifth-most populous jurisdiction). So Baltimore City is over-performing expectations, but otherwise the pattern mostly follows the population.

```{r}
new_voters <- md_voters_2024 |> 
  filter(STATE_REGISTRATION_DATE > '2024-07-21') |> 
  filter(STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE)

new_by_county <- new_voters |> 
  group_by(COUNTY) |> 
  summarize(voters = n()) |> 
  arrange(desc(voters))
```

**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2** The two full weeks after July 21 saw the second and third-highest number of new registrants, only surpassed by week 24, which was in mid-June. So some evidence that weeks 31 and 32 are outliers.

```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(
    week = week(COUNTY_REGISTRATION_DATE)
  )

md_voters_2024 |>
  group_by(week) |>
  summarise(
    total = n()
  ) |>
  arrange(desc(total))
```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** July, with 26,284, so seems like more evidence of a larger than usual bump.

```{r}
md_voters_2024 |>
  mutate(month = floor_date(COUNTY_REGISTRATION_DATE, "month")) |>
  group_by(month) |>
  summarise(total = n()) |>
  arrange(desc(total))
```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** The Capital and Central regions dominate new registrations, which makes sense since they are the most populated portions of the state. In those regions, Democrats and unaffiliated new voters vastly outnumber Republicans. In Southern Maryland, unaffiliated new voters lead both parties.

```{r}
md_voters_2024 <- md_voters_2024 |> 
  mutate(region = case_when(
    COUNTY == 'Allegany' ~ 'Western',
    COUNTY == 'Garrett' ~ 'Western',
    COUNTY == 'Washington' ~ 'Western',
    COUNTY == 'Frederick' ~ 'Capital',
    COUNTY == 'Montgomery' ~ 'Capital',
    COUNTY == "Prince George's" ~ 'Capital',
    COUNTY == 'Anne Arundel' ~ 'Central',
    COUNTY == 'Baltimore City' ~ 'Central',
    COUNTY == 'Baltimore County' ~ 'Central',
    COUNTY == 'Carroll' ~ 'Central',
    COUNTY == 'Harford' ~ 'Central',
    COUNTY == 'Howard' ~ 'Central',
    COUNTY == 'Calvert' ~ 'Southern',
    COUNTY == 'Charles' ~ 'Southern',
    COUNTY == "Saint Mary's" ~ 'Southern',
    COUNTY == 'Kent' ~ 'Eastern Shore',
    COUNTY == "Queen Anne's" ~ 'Eastern Shore',
    COUNTY == 'Talbot' ~ 'Eastern Shore',
    COUNTY == 'Caroline' ~ 'Eastern Shore',
    COUNTY == 'Dorchester' ~ 'Eastern Shore',
    COUNTY == 'Wicomico' ~ 'Eastern Shore',
    COUNTY == 'Somerset' ~ 'Eastern Shore',
    COUNTY == 'Worcester' ~ 'Eastern Shore',
    COUNTY == 'Cecil' ~ 'Eastern Shore'
  ))

md_voters_2024 |> 
  group_by(region, PARTY) |> 
  summarize(voters = n()) |> 
  arrange(desc(voters))

```

**Q5** What's an additional question you could ask of this data?

**A5** 
