---
title: "lab_02"
author: "derek willis"
date: "8/09/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the slice of WinRed contribution data from West Virginia.  

```{r}
wv_winred_contribs <- read_rds("data/lab_02.rds")
```

## Examine Data

## Answer questions

**Q1** Create a new column that divides the contribution amounts into the following options - less than $10, $10-$25, $26-$50, $51_$100, $101-$500, $501 and up - and calculates how many contributions are in each group. How would you describe the distribution of contributions in a sentence?
A1. The two largest groups are for contributions of $25 or less, with only a few dozen contributions that are larger than $500.

```{r}
wv_winred_contribs %>%
  mutate(
    amount_group = case_when(
      amount < 10 ~ 'under_10',
      amount >= 10 & amount <= 25 ~ '10_25',
      amount > 25 & amount <= 50 ~ '26_50',
      amount > 50 & amount <= 100 ~ '51_100',
      amount > 100 & amount <= 500 ~ '101_500',
      amount > 500 ~ '501_plus'
    )
  ) %>%
  group_by(amount_group) %>%
  summarise(
    total = n()
  ) %>%
  arrange(desc(total))
```

Q2. Which West Virginia donors made the most contributions of less than $10? You'll need to use a combination of columns to get unique donors.  
A2. The Rev. Dr. John Almond of Beckley made at least 293 contributions of less than $10, more than 100 more than the next most-frequent donor, Franco Colonna of South Charleston.

```{r}
wv_winred_contribs %>%
  filter(amount < 10) %>% 
  group_by(last_name, first_name, address_one, city) %>%
  summarise(
    total = n()
  ) %>%
  arrange(desc(total))
```

Q3. What questions do the results of Q2 raise in your mind? How could you use the data to explore them?
A3. Potential questions might include who Almond and Colonna give to and how frequently, or why a pastor gives political contributions (there's no prohibition, but many clergy avoid it). Is there a pattern of giving among the most frequent donors? We could use the dates, recipients and amounts to help explore these questions.

Q4. Which month saw the most contributions?
A4. June 2022, with 4,636.

```{r}
wv_winred_contribs %>%
  mutate(Month = floor_date(date, "month")) %>%
  group_by(Month) %>%
  summarise(total = n()) %>%
  arrange(desc(total))
```

Q5. How many contributions of $2,900 - the individual maximum donation - were there, and who did they go to? How many of the recipients are from West Virginia? Do some googling to find out.
A5. There are 6 contributions of $2,900, and three of them went to Carol Miller, while one each went to David McKinley, Ryan Zinke and Steven Palazzo. Miller and McKinley are from WV.

```{r}
wv_winred_contribs %>%
  filter(amount == 2900) %>%
  arrange(date)
```
