---
title: "lab_02"
author: "derek willis"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with West Virginia loans contained in the PPP loans data.  

```{r}
west_virginia_ppp <- read_rds("data/lab_02.rds")
```

## Examine Data

## Answer questions

**Q1** Create a new column that divides the loan amounts into the following options - under_10k, 10k_50k, 50k_100k, 100k_500k, 500k_1m, 1m_plus - and calculates how many loans are in each group. How would you describe the distribution of loans in a sentence?
A1. The 10k-50k group has 15,089, the most of any group. More than 2/3rds of loans are for less than $50,000, with only a few hundred over $1 million, suggesting that the program reached actual small businesses.

```{r}
west_virginia_ppp %>%
  mutate(
    amount_option = case_when(
      amount < 10000 ~ 'under_10k',
      amount >= 10000 & amount < 50000 ~ '10k_50k',
      amount >= 50000 & amount < 100000 ~ '50k_100k',
      amount >= 100000 & amount < 500000 ~ '100k_500k',
      amount >= 500000 & amount < 1000000 ~ '500k_1m',
      amount >= 1000000 ~ '1m_plus'
    )
  ) %>%
  group_by(amount_option) %>%
  summarise(
    total_loans = n()
  ) %>%
  arrange(desc(total_loans))
```

Q2. Which servicing lenders have the most loans with undisbursed amounts greater than 0?
A2. Davis Trust Company tops the list with 12.

```{r}
west_virginia_ppp %>%
  filter(undisbursed_amount > 0) %>%
  group_by(servicing_lender_name, servicing_lender_state) %>%
  summarise(
    total_loans = n()
  ) %>%
  arrange(desc(total_loans))
```

Q3. What are some potential reasons why a loan might not be paid out? What questions does that circumstance raise for you?
A3. An applicant might apply not knowing if they needed the loan and then decline it. Or the lender might decide not to fund the loan for some reason - lack of information or something suspicious. Lots of potential questions here, including how often this has happened and whether undisbursed loans have some patterns.

Q4. Which month in 2021 had the most loans approved?
A4. March 2021, with 4,394.

```{r}
west_virginia_ppp %>%
  filter(date_approved > '2020-12-31') %>%
  mutate(Month = floor_date(date_approved, "month")) %>%
  group_by(Month) %>%
  summarise(total = n()) %>%
  arrange(desc(total))
```

Q5. How many loans went to West Virginia applicants that listed the maximum number of jobs retained (500)? Which one had the earliest approved date?
A5. There were 10 applicants that listed 500 jobs retained. J. AUSTIN HEATING & COOLING, INC, which was approved on April 8, 2020, had the earliest approved date.

```{r}
west_virginia_ppp %>%
  filter(jobs_retained == 500) %>%
  arrange(date_approved) %>%
  select(name, amount, date_approved)
```
