---
title: "lab_12"
author: "Derek Willis"
date: "2024-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* tidytext and our usual libraries

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}

```

## Questions

**Q1.** You've been assigned to report a story about the leading reasons that Maryland attorneys get sanctioned by the state for misconduct. The state [publishes lists of sanctions](https://www.courts.state.md.us/attygrievance/sanctions) that contain a short text description about the situation. Load the CSV file in the data folder containing records from fiscal year 2011 onwards. Make a list of unique words from the text column, then following the example in the pre_lab, remove common "stop words" from that list and create a list of the top 25 words containing the percentage of occurrences each word represents. What's the leading word in that answer and, broadly, what do you think the top 25 words describe?

**A1.** 

```{r}
md_attorney_sanctions <- read_csv("data/md_attorney_sanctions.csv")

unique_words <- md_attorney_sanctions |> select(text) |>
  unnest_tokens(word, text)

data("stop_words")

unique_words |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(25)

```

**Q2.** Let's move beyond single words to phrases. Make a list of the top 10 three-word phrases, called trigrams, based on the example from the pre_lab (you'll need to modify the example code to do this). What's the top trigram and how often does it appear? What does that phrase mean in legal terms?

**A2.** The top phrase is "attorney trust account," which appears 343 times and represents 5.8 percent of all trigrams. Attorney trust accounts are bank accounts where clients place money that is held or controlled by the attorney, making it a possible vector for misuse.

```{r}
md_attorney_sanctions |>
  unnest_tokens(fiver, text, token = "ngrams", n = 5) |>
  separate(fiver, c("word1", "word2", "word3", "word4", "word5"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  filter(!word3 %in% stop_words$word) |>
  filter(!word4 %in% stop_words$word) |>
  filter(!word5 %in% stop_words$word) |>
  mutate(fiver = paste(word1, word2, word3, word4, word5, sep=" ")) |>
  group_by(fiver) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)
```

**Q3.** Let's drop back down to more traditional analysis - take the top trigram from Q2 and write code to see how many times it occurs in the text column in each fiscal year. If you come across an error (you may not), google the error message and tack on "tidytext". What do you think the answer produced by your code suggests?

**A3.** FY 2021 saw the most sanction actions mentioning "attorney trust account", with 37, slightly more than in 2022, which had 32. Since the start of the pandemic each year has seen totals that exceed any pre-pandemic year (and there are just three so far in FY 2023, which suggests that things are returning to a more normal level).

```{r}
md_attorney_sanctions |> 
  filter(str_detect(text, "attorney trust account")) |> 
  group_by(fiscal_year) |> 
  tally(sort=TRUE)


```
