# Geographic analysis

In the previous chapter, we looked at election precincts in Prince George's County to show a bit of a pattern regarding concentration of the precincts with the most and 0 eligible voters. Let's go little further and look at voters statewide.

First, let's load the libraries we'll need. We're also going to load tidycensus and set an API key for tidycensus.

```{r}
#| output: false
library(tidyverse)
library(sf)
library(janitor)
library(tidycensus)
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

```

For the rest of this chapter, we're going to work on building a map that will help us gain insight into geographic patterns in voter registration by county in Maryland. Our question: by examining the number of Democrats/Republicans/Unaffiliated voters per 100,000 people in each county, what regional geographic patterns can we identify?

We've got voters by county, so let's load that and take a look:

```{r}
voters_by_county <- read_csv("data/eligible_voters.csv")

voters_by_county %>% arrange(desc(TOTAL))
```

So, what do we see here? Montgomery County has the most, followed by Prince George's & Baltimore County. Checks out.

Next, we'll go out and get population data for each county from tidycensus. The variable for total population is B01001_001.  

```{r}
md_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              state = "MD")

md_county_population
```

Ultimately, we're going to join this county population table with our voters by county table, and then calculate a voters per 50,000 people statistic.  But remember, we then want to visualize this data by drawing a county map that helps us pick out trends. Thinking ahead, we know we'll need a county map shapefile.  Fortunately, we can pull this geometry information right from tidycensus at the same time that we pull in the population data by adding "geometry = TRUE" to our get_acs function.

```{r}
md_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              state = "MD",
              geometry = TRUE)

md_county_population
```

We now have a new column, geometry, that contains the "MULTIPOLYGON" data that will draw an outline of each county when we go to draw a map.

The next step will be to join our population data to our voter data on the county column.

But there's a problem.  The column in our population data that has county names is called "NAME", and it has the full name of the county spelled out in title case -- first word capitalized and has "County" and "Maryland" in it.  The voter data just has the name of the county.  For example, the population data has "Anne Arundel County, Maryland" and the voter data has "Anne Arundel".

```{r}
md_county_population

voters_by_county
```

If they're going to join properly, we need to clean one of them up to make it match the other.  

Let's clean the population table. We're going to rename the "NAME" column to "County", then remove ", Maryland" and "County" and make the county titlecase. Next we'll remove any white spaces after that first cleaning step that, if left in, would prevent a proper join. We're also going to rename the column that contains the population information from "estimate" to "population" and select only the county name and the population columns, along with the geometry. That leaves us with this tidy table.

```{r}
md_county_population <- md_county_population %>%
  rename(County = NAME) %>%
  mutate(County = str_to_title(str_remove_all(County,", Maryland|County"))) %>%
  mutate(County = str_trim(County,side="both")) %>%
  rename(population = estimate) %>%
  select(County, population, geometry)

md_county_population
```

Now we can join them.  

```{r}
md_voters_per_10k <- md_county_population %>%
  left_join(voters_by_county)

md_voters_per_10k
```

Hang on - there's at least one county with NA values - St. Mary's, which is spelled "Saint Mary's" in the voter dataframe. And Baltimore County didn't match, either. Let's fix that using if_else, which allows us to conditionally mutate:

```{r}
md_county_population <- md_county_population %>%
  mutate(County = if_else(County == "St. Mary's", "Saint Mary's", County)) %>% 
  mutate(County = if_else(County == "Baltimore", "Baltimore County", County))
```

Our final step before visualization, let's calculate the number of voters per 10,000 population for each county and sort from highest to lowest to see what trends we can identify just from the table.

```{r}
md_voters_per_10k <- md_county_population %>%
  left_join(voters_by_county) %>%
  mutate(voters_per_10k = TOTAL/population*10000) %>%
  arrange(desc(voters_per_10k))

md_voters_per_10k
```

Let's take a look at the result of this table.  There are some surprising ones at the top, some of Maryland's smallest counties! Worcester, Queen Anne's, Talbot may not have that many voters, but they also don't have a lot of people.  

Okay, now let's visualize.  We're going to build a choropleth map, with the color of each county -- the fill -- set according to the number of voters per 10K on a color gradient.

```{r}
ggplot() +
  geom_sf(data=md_voters_per_10k, aes(fill=voters_per_10k)) +
  theme_minimal()
```

This map is okay, but the color scale makes it hard to draw fine-grained differences. Let's try applying the magma color scale we learned in the last chapter.

```{r}
ggplot() +
  geom_sf(data=md_voters_per_10k, aes(fill=voters_per_10k)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma")
```

The highest ranking counties stand out nicely in this version, but it's still hard to make out fine-grained differences between other counties.

So let's change the color scale to a "log" scale, which will help us see those differences a bit more clearly.

```{r}
ggplot() +
  geom_sf(data=md_voters_per_10k, aes(fill=voters_per_10k)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log")
```

Let's repeat that for Unaffiliated voters:

```{r}
md_voters_per_10k <- md_voters_per_10k %>% 
  mutate(una_voters_per_10k = UNA/population*10000)
```

And then map it:

```{r}
ggplot() +
  geom_sf(data=md_voters_per_10k, aes(fill=una_voters_per_10k)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log")
```

What regional patterns do you see, especially on the ends of the scale?