---
title: "pre_lab_04.Rmd"
author: "derek willis"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Points to hit
1. Review of third lab questions/problems.
2. Demonstration of combining and merging

## Chapter 10

### Task 1: Load libraries and settings
**Task** Run the following code in the gray-colored codeblock below to load the tidyverse library and turn off scientific notation.  

```{r}
# turn off sci notation
options(scipen=999)
library(tidyverse)
library(lubridate)
```

### Task 2: Load data
Let's say that we have Maryland county voter registration data from three different elections in three different files. They have the same record layout and the same number of counties (plus Baltimore City).

**Task** Load the three CSV population estimate files. Do they have the same column structure?
**Answer**

```{r}
county_voters_2016 <- read_csv("data/county_voters_2016.csv")
```

```{r}
county_voters_2018 <- read_csv("data/county_voters_2018.csv")
```

```{r}
county_voters_2020 <- read_csv("data/county_voters_2020.csv")
```


### Task 3: Using bind_rows
**Task** Combine the three files into a single data frame using the bind_rows function. Add a description of what this code does to your reference notebook.

```{r}
# bind_rows with list
county_voters_combined <- bind_rows(list(county_voters_2016, county_voters_2018, county_voters_2020))
View(county_voters_combined)
```

## Joining data

### Task 4: Loading population data
**Task** Load the Maryland 2020 county population data
**Answer**

```{r}
maryland_population <- read_csv('data/maryland_population_2020.csv')
```

### Task 5: Join population and voter data
**Task** Run the following code to join Maryland population data and the 2020 voter registration data together using the common column county as the key. How many rows are there now? How many *should* there be?
**Answer**

```{r}
# with nrow included to show row total
maryland_population %>% left_join(county_voters_2020, by="COUNTY") %>% nrow()
```
### Task 6: Looking at Joined Data
**Task** Examine the combined data

```{r}
# without nrow
maryland_population %>% left_join(county_voters_2020, by="COUNTY")
```

### Task 7: Assign Joined Data to New Dataframe
**Task** Create a new dataframe with PPP and NAICS data joined together. Add a description of what this code does to your reference notebook.

```{r}
maryland_population_with_voters <- maryland_population %>% left_join(county_voters_2020)
```

### Task 8: Add the Demographic Data
**Task** Load the Maryland demographic data

```{r}
maryland_demographics <- read_csv('data/maryland_demographics.csv')
```

### Task 9: Join demographic data to combined voting/population data
**Task** Join the ZCTA data to the our PPP & NAICS data. What's the error message you get? What do you think it means? Add a description of what this code does to your reference notebook.
**Answer**

```{r}
# join on zip - but there's a problem
maryland_population_with_voters_and_demographics <- maryland_population_with_voters %>% left_join(maryland_demographics, by=c("COUNTY"="NAME"))
```

### Task 10: Asking Demographic Questions
**Task** Using mutate, let's find the county with the highest 18+ Black population as a percentage of all population 18 and over and also the percentage of Democrats in that county. Put your answers below.
**Answer**

```{r}
maryland_population_with_voters_and_demographics %>%
  mutate(pct_black_18_plus = (pop_black/pop_18_over)*100, pct_dems = (DEM/Total)*100) %>% 
  arrange(desc(pct_black_18_plus)) %>% 
  select(COUNTY, pct_black_18_plus, pct_dems)
```


### Task 11: Asking Demographic Questions
**Task** Now let's find the county with the highest 18+ Asian population as a percentage of all population 18 and over and also the percentage of Democrats in that county. How do these results compare to the previous task? Put your answers below.
**Answer**

```{r}
maryland_population_with_voters_and_demographics %>%
  mutate(pct_asian_18_plus = (pop_asian/pop_18_over)*100, pct_dems = (DEM/Total)*100) %>% 
  arrange(desc(pct_asian_18_plus)) %>% 
  select(COUNTY, pct_asian_18_plus, pct_dems)
```

### Task 15: Asking Demographic Questions
**Task** Come up with 2 other questions that you could ask of this data, with the caveat that they can't be variations of the ones above.
**Answer**


