---
title: "lab_03"
author: "sean mussenden"
date: "8/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data
```{r}
# Load WV slice of ppp loan data
wv_ppp_loans <- read_rds("lab_03.rds")
```

## Answer questions

Q1: Which five cities located in Kanawha County, West Virginia, got the largest _number_ of loans? How many did each get? To answer this question, you'll have to clean the very messy "city" field.

Note: To answer this question, you'll need to see every variation of the "city" column in Kanawha County, and then fully clean the city field for those cities. Hint: once cleaned, the dataframe that counts the number of loans by city should have 38 distinct cities for Kanawha County.

A1: The top five cities located in Kanawha County that got loans are Charleston (2837), South Charleston (484), Saint Albans (304), Dunbar (166), Nitro (142).  

```{r}
kanawha_ppp_loans_by_city <- wv_ppp_loans %>%
  filter(project_county_name == "KANAWHA") %>%
  mutate(city=str_to_title(city)) %>%
  mutate(city=case_when(
    str_detect(city, "^St") ~ "Saint Albans",
    str_detect(city,"^S |^So ") ~ str_replace(city, "^S |^So ", "South "),
    city == "Balle" ~ "Belle",
    city == "Charelston" ~ "Charleston",
    city == "Crosslanes" ~ "Cross Lanes",
    city == "Elkveiw" ~ "Elkview",
    TRUE ~ city
  )) %>%
  group_by(city) %>%
  summarise(total_loans = n()) %>%
  arrange(desc(total_loans))

kanawha_ppp_loans_by_city

```

Q2: Find the 5-digit ZIP code in West Virginia that had the highest total amount of loans (by dollar value). What is the name of the business that got the largest loan (by dollar value) in that zip code? Write up a short paragraph that also answers these questions:

* Where in West Virginia is this ZIP code located?
* How much was the business' loan for?
* How much of that loan was forgiven?
* Which number is bigger, the loan amount or the forgiveness amount? Can you explain the discrepancy?

A1: A ZIP code (25301) in Charleston, West Virginia's capital, got more PPP loan money than anywhere else in the state, and no business in that ZIP code got more money in a single loan than Jackson Kelly, a large law firm that has represented opioid distributors and has offices across Appalachia.

Jackson Kelly got a total of \$5.07 million, which it said it used to retain 225 jobs.  It had \$5.13 million forgiven, which is more than the original loan amount.  The most likely explanation, one that would require reporting to confirm, is that the $6,000 discrepancy represents the interest Jackson Kelly had paid on the loan in the year between when it was issued and when it was forgiven.

```{r}
# Create a dataframe that shows the top ZIP code by total dollar amount
wv_top_zip_code <- wv_ppp_loans %>%
  mutate(zip_five = str_sub(zip, start=1L, end=5L)) %>%
  group_by(zip_five) %>%
  summarise(
    total_amount = sum(amount)
  ) %>%
  arrange(desc(total_amount))

wv_top_zip_code

# Use the top ZIP code in the dataframe you just created and create a new dataframe that shows the business with the biggest loan in that ZIP code.
wv_top_biz_in_top_zip <- wv_ppp_loans %>%
  mutate(zip_five = str_sub(zip, start=1L, end=5L)) %>%
  filter(zip_five == "25301") %>%
  arrange(desc(amount)) %>%
  head(1)

wv_top_biz_in_top_zip

```

Q3: Five business entities owned by the spouse of a member of West Virginia's U.S. House delegation got PPP loans. Write up a paragraph that answers the following questions.   

  * What are the businesses' names?
  * What do the businesses do?
  * How much money did the five businesses receive in total?
  * What percentage of the total loan amount have been forgiven?
  * How did the U.S. House member vote on the legislation initially authorizing the PPP loan program.

Feel free to add any interesting "color" (little interesting details) you find in your research.  

Some hints to help you answer this question:

* West Virginia's three U.S. House members are listed [here](https://www.govtrack.us/congress/members/WV#representatives)
* To find businesses they or their families own, you can search for financial disclosure forms for U.S. House members [here](https://disclosures-clerk.house.gov/PublicDisclosure/FinancialDisclosure#Search).
  * To speed things up, here are direct links to the three U.S. House members most recent financial disclosures, which I found by searching the site linked above.
  * [Rep. 1](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10042042.pdf)
  * [Rep. 2](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10040330.pdf)
  * [Rep. 3](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10042734.pdf)
* When you examine the disclosure forms, there may be lots of stock sales and retirement accounts mixed in.  Ignore those. In the table in the disclosure forms, look for "Partnership" or "C-Corp" in the column named "Income Type"
* This is a link to details on passage of the [CARES ACT](https://www.congress.gov/bill/116th-congress/house-bill/748). You can find a link to how the House voted by clicking on a link from [this page](https://www.congress.gov/bill/116th-congress/house-bill/748/actions).

A3: Rep. Carol Miller, who represents West Virginia's 3rd District, voted for the legislation that initially authorized the PPP loan program, a program that directed millions of dollars to her family's auto dealership businesses, Dutch Miller. Company motto: ["If You Don't Buy Dutch You Will Pay Way Too Much!"](https://www.dutchmillerauto.com/aboutus.aspx).  Her official House bio conveniently avoids the family's auto business, though, [calling her a bison farm operator and real estate manager](https://miller.house.gov/about). The business is run by her husband and son's, according to the company's website. Five separate legal entities that are part of the family auto dealership empire received loans, totaling \$3.11 million.  

* DM MOTOR, INC
*	DUTCH MILLER OF CHARLESTON, INC
* DUTCH MILLER OF CHARLOTTE, INC
* DUTCH MILLER SUBARU, INC
* DUTCH MILLER CHEVROLET, INC.

The loans were entirely forgiven, plus extra, most likely to account for interest paid.  So, in effect, Miller voted for a program that paid her family \$3 million! Pretty interesting peg for a story. [Lots of members of Congress got loans for family businesses](https://www.usatoday.com/story/news/politics/2020/07/06/ppp-loans-firms-tied-lawmakers-among-those-getting-pandemic-aid/5382594002/).

```{r}

# Create a dataframe of loans to companies
miller_ppp_loans <- wv_ppp_loans %>%
  filter(str_detect(name,"DUTCH |DM MOTOR"))

miller_ppp_loans

# Calculate total loan amount and percentage forgiven for those companies
miller_totals <- miller_ppp_loans %>%
  mutate(amount = as.numeric(amount)) %>%
  summarise(
    total_amount = sum(amount),
    total_forgiveness_amount = sum(forgiveness_amount)
  ) %>%
  mutate(pct_forgiven = (total_forgiveness_amount/total_amount)*100)

miller_totals

```
