---
title: "lab_03"
author: "Derek Willis"
date: "2024-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds")
```

## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

A1: The top seven cities are BETHESDA (1177), CHEVY CHASE (704), BALTIMORE (690), ANNAPOLIS (490), POTOMAC (490), SILVER SPRING (405) and ROCKVILLE (358).

```{r}
# initial list of cities are: BALTIMORE, BETHESDA, SILVER SPRING, CHEVY CHASE, POTOMAC, ROCKVILLE, ANNAPOLIS
md_contribs_by_city <- md_contribs |>
  group_by(contributor_city) |> 
  summarize(count = n()) |> 
  arrange(desc(count))

# clean those cities
md_contribs_by_city_fixed <- md_contribs |> 
  mutate(contributor_city=case_when(
    contributor_city == "BALRIMORE" ~ "BALTIMORE",
    contributor_city == "ANAPOLIS" ~ "ANNAPOLIS",
    contributor_city == "CHEVY CHASE M.D." ~ "CHEVY CHASE",
    contributor_city == "POTAMIC" ~ "POTOMAC",
    contributor_city == "SILVER SPRINGS" ~ "SILVER SPRING",
    TRUE ~ contributor_city
  )) |>
  group_by(contributor_city) |>
  summarise(total = n()) |>
  arrange(desc(total))

md_contribs_by_city_fixed

```

Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). Where is that zip code and what kind of area is it? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

A2: A ZIP code (20817) in Bethesda is responsible for more contributions to candidates this year than anywhere else in the state. Bethesda is a very wealthy area bordering D.C. and home to prominent business people, government officials and others with a stake in government.

Dan Helmer, a Democrat running in a district just across the Potomac River in Virginia, was the leading recipient. Just two of the top five are Maryland candidates (Democrats Jamie Raskin and April McClain Delaney).

```{r}
# Create a dataframe that shows the top ZIP code by total dollar amount
md_contribs_zip <- md_contribs |>
  mutate(zip_five = str_sub(contributor_zip, start=1L, end=5L)) |>
  group_by(zip_five) |>
  summarise(
    total_amount = sum(contribution_receipt_amount)
  ) |>
  arrange(desc(total_amount))

md_contribs_zip

# Use the top ZIP code in the dataframe you just created and create a new dataframe that shows the committee that got the most money from that ZIP code.
md_zip_top_recipient <- md_contribs |>
  mutate(zip_five = str_sub(contributor_zip, start=1L, end=5L)) |>
  filter(zip_five == "20817") |>
  group_by(committee_name) |>
  summarise(
    total_amount = sum(contribution_receipt_amount)
  ) |>
  arrange(desc(total_amount))
  head(1)

md_zip_top_recipient

```

Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

A3: The top individual Maryland donor to U.S. House campaigns is Melanie Jacobsen, a former SEC senior special counsel, and her spouse Mark Jacobsen. They describe their occupations as farming and banking, respectively. Looking at the top ten, Mark appears twice, so he actually might be the top donor. William and Bill Pugh might also be the same person.

```{r}

# Create a dataframe of top individual contributors
md_contribs_top_indiv <- md_contribs |>
  group_by(contributor_name, contributor_city) |> 
  summarize(total_amount = sum(contribution_receipt_amount)) |> 
  arrange(desc(total_amount))

md_contribs_top_indiv

# Check out Jacobsen donations
jacobsen_donations <- md_contribs |>
  filter(str_detect(contributor_name, '^JACOBSEN,'))
```
