---
title: "lab_02"
author: "Derek Willis, adapted by Aidan Hughes"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters.

Which counties have seen the most? Take a look at county population figures here and compare that list with your data findings in your answer: https://worldpopulationreview.com/us-counties/maryland.

**A1**

```{r}
new_voters_post_biden <- md_voters_2024 |>
  filter(STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE) |>
  filter(STATE_REGISTRATION_DATE > "2024-07-21")

new_voters_post_biden
```

Over 17,000 newly registered voters after Biden dropped out.


```{r}
new_voters_post_biden |>
  group_by(COUNTY) |>
  summarise(count = n()) |>
  arrange(desc(count))
```

Montgomery, Baltimore City, Baltimore County, Prince George's, and Anne Arundel had the most new voters. As a group, those are the most populous counties in MD, but they're not quite in order of population


**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks that year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2**

```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(week_registered = week(STATE_REGISTRATION_DATE))
```

```{r}
new_voters_weekly <- md_voters_2024 |>
  filter(STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE) |>
  group_by(week_registered) |>
  summarise(number_registered = n()) |>
  arrange(desc(number_registered))

new_voters_weekly
```

Those two weeks after Biden stepped back saw the 2nd and 3rd largest number of new registrations of any week in our data. However, another week--week 24, ending around June 14--saw an even higher number.


**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?

**A3** 

```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(month_registered = month(STATE_REGISTRATION_DATE))
```

```{r}
monthly_registrations <- md_voters_2024 |>
  group_by(month_registered) |>
  summarise(count = n()) |>
  arrange(desc(count))

monthly_registrations
```
July had by far the largest number of registrations, followed by June. That aligns with those two weeks after Biden stepped back being outliers vs normal rates of registration.

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are.

Which regions and parties account for the most new voters?

**A4**

```{r}
md_voters_2024 <- md_voters_2024 |> 
  mutate(region = case_when(
    COUNTY == 'Allegany' ~ 'Western',
    COUNTY == 'Garrett' ~ 'Western',
    COUNTY == 'Washington' ~ 'Western',
    COUNTY == 'Frederick' ~ 'Capital',
    COUNTY == 'Montgomery' ~ 'Capital',
    COUNTY == "Prince George's" ~ 'Capital',
    COUNTY == 'Anne Arundel' ~ 'Central',
    COUNTY == 'Baltimore City' ~ 'Central',
    COUNTY == 'Baltimore County' ~ 'Central',
    COUNTY == 'Carroll' ~ 'Central',
    COUNTY == 'Harford' ~ 'Central',
    COUNTY == 'Howard' ~ 'Central',
    COUNTY == 'Calvert' ~ 'Southern',
    COUNTY == 'Charles' ~ 'Southern',
    COUNTY == "Saint Mary's" ~ 'Southern',
    COUNTY == 'Kent' ~ 'Eastern Shore',
    COUNTY == "Queen Anne's" ~ 'Eastern Shore',
    COUNTY == 'Talbot' ~ 'Eastern Shore',
    COUNTY == 'Caroline' ~ 'Eastern Shore',
    COUNTY == 'Dorchester' ~ 'Eastern Shore',
    COUNTY == 'Wicomico' ~ 'Eastern Shore',
    COUNTY == 'Somerset' ~ 'Eastern Shore',
    COUNTY == 'Worcester' ~ 'Eastern Shore',
    COUNTY == 'Cecil' ~ 'Eastern Shore'
  ))
```


you could do the next part in either two ways based on how you interpreted the question:

1) do separate group_by/summarizes, one for region and one for PARTY
2) do a group_by with both region and PARTY together, and then do your summarize

here's number 1):

```{r}
md_voters_2024 |>
  group_by(region) |>
  summarise(count = n()) |>
  arrange(desc(count))
```
```{r}
md_voters_2024 |>
  group_by(PARTY) |>
  summarise(count = n()) |>
  arrange(desc(count))
```

Here's number 2):

```{r}
md_voters_2024 |>
  group_by(PARTY, region) |>
  summarise(count = n()) |>
  arrange(desc(count))
```


**Q5** What's an additional question you could ask of this data?


**A5** 
