---
title: "lab_11"
author: "sean mussenden"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Census API key

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
library(janitor)
library(sf)
library(tigris)

# Establish API Key for Census
census_api_key("549950d36c22ff16455fe196bbbd01d63cfbe6cf")

```

## Load data
**Task** Create a codeblock and load the WV slice of the PPP loan data, which is in the data folder and is called lab_11.rds.

```{r}
wv_ppp <- read_rds("data/lab_11.rds")

```

## Questions 

**Q1.** You are interested in doing some reporting on how West Virginia companies in the "Mining, Quarrying, and Oil and Gas Extraction" industry (https://www.naics.com/six-digit-naics/?code=21) used PPP loans during the pandemic (West Virginia is a big coal mining state).  

Then produce a choropleth map that shows West Virginia counties shaded according to the total amount of loans to that industry per 100,000 population. 

Use the magma color palette, and shade on a "log" scale. Also produce a table that shows the total amount of loans per 100,000 population for each county, ranked from highest to lowest, along with the total amount of loans and the total number of loans. 

Then answer these questions:

* Based on the map, what region of the state appears to have gotten more loans in this industry? What geographic pattern do you notice in the data?
* Based on the table, what county got the highest total loan amount? What was the total amount per 100K population, the total amount of loans (not population adjusted), how many loans were there, and what's the population?

**A1.** There's a southwest-to-northeast seam running through the center of the state.  The counties on West Virginia's east and west sides have lower loan amounts (and some have no loans at all, in grey). Ritchie County got the highest amount per 100K: $77 million dollars.  There were 9 total loans of 7.65 million. The county has a population of 9844. 
```{r}

# Create a table with WV mining industries only, group by county and calculate total loans and total amount.
wv_mining <- ppp_wv_loans |>
  mutate(two_digit_naics = str_sub(naics_code,start=1L, end=2L)) |>
  filter(two_digit_naics == "21") |>
  group_by(project_county_name) |>
  summarise(
    total_loans = n(),
    total_amount = sum(amount)
  )
  
# Get a population table from tidycensus and clean it to prepare to join
wv_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              state = "WV",
              geometry = TRUE) |>
  rename(COUNTY = NAME) |>
  mutate(COUNTY = toupper(str_remove_all(COUNTY,", West Virginia|County"))) |>
  mutate(COUNTY = str_trim(COUNTY,side="both")) |>
  rename(population = estimate) |>
  select(COUNTY, population, geometry) |>
  clean_names()

# Join mining to population and calculate amount per 100k, also replace any nas with 0s.
wv_mining_population <- wv_county_population |>
  left_join(wv_mining,by=c("county" = "project_county_name")) |>
  mutate(amount_per_100k = total_amount/population*100000) |>
  mutate(amount_per_100k = replace_na(amount_per_100k,0)) 


# Plot it on a log scale, shading according to amount per 100k
ggplot() +
  geom_sf(data=wv_mining_population, aes(fill=amount_per_100k)) +
  theme_minimal() + 
  scale_fill_viridis_b(option="magma", trans="log")

# Output a table sorted by amount_per_100k
wv_mining_population |>
  as_tibble() |>
  arrange(desc(amount_per_100k))


```

**Q2.** You want to understand WHY the spatial pattern you saw above exists.  So you grab a shapefile with the latitude and longitude coordinates of every mine permit in West Virginia from the state Department of Environmental Protection. 

Here's the url: https://tagis.dep.wv.gov/arcgis/rest/services/WVDEP_enterprise/mining_reclamation/MapServer/0/query?where=1%3D1&outFields=*&geometry=&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=4326&f=json. 

(It's in a slightly different format than the .shp file in the lab. It's another type of shapefile called "geojson".  It will work just the same, though, despite that difference. You can read it directly into R without manually downloading it, by putting the url in the st_read() function we learned about in this week's prelab.) 

This data includes both currently operating mines and old ones that have closed.  You want to examine only mines that have a permit expiration date after Sept. 1, 2021. 

Make a map with a point for each one of these mine permits and put a blank map with the outline of West Virginia counties (which you can get from Tigris) behind it. 

What spatial pattern do you see? How does it compare to the spatial pattern in the first map you made?

**A2.** It's pretty similar.  There's a big cluster of mines in the southern part of the state, and most of them are in the southwest-northeast seam we identified previously.

```{r}

# Read in the data
wv_mine_locations <-st_read("https://tagis.dep.wv.gov/arcgis/rest/services/WVDEP_enterprise/mining_reclamation/MapServer/0/query?where=1%3D1&outFields=*&geometry=&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=4326&f=json")

# Filter for only active mines based on expire date
wv_active_mine_locations <- wv_mine_locations  |>
  mutate(expiredate = ymd(expiredate)) |>
  filter(expiredate > "2021-09-01")

# Get a counties table from Tigris
wv_counties <- counties() |>
  filter(STATEFP == "54")

# Make a map
ggplot() +
  geom_sf(data=wv_counties) +
  geom_sf(data=wv_active_mine_locations) +
  theme_minimal()

```
Q3. To confirm the spatial comparison you identified in question 2, let's create a new map that layers the mine location points on top of our PPP choropleth map we created in question 1. What do you notice about the amount of mines in the county that has the highest total loan amount per 100K?

A3. The county (the only one in yellow) doesn't appear to be in a part of West Virginia with a lot of mines. 

```{r}

ggplot() +
  geom_sf(data=wv_mining_population, aes(fill=amount_per_100k)) +
  geom_sf(data=wv_active_mine_locations) +
  theme_minimal() + 
  scale_fill_viridis_b(option="magma", trans="log")

```
Q4. Starting from the original WV ppp dataframe, examine the loans to companies in "Mining, Quarrying, and Oil and Gas Extraction" in the county with the highest total loan amount per 100K (you can use the table you created in question 1 to identify the top county). Answer the following questions:

* What single company accounts for the biggest chunk of the total loan amount? 
* What specific industry are they in, based on their 6-digit NAICS code? Look them up on the web. What services do they provide?
* How many jobs did they save, according to the data?
* What's your hypothesis for why a company in their specific industry would need to seek a PPP loan during the pandemic?

A4. Hall Drilling, Inc. (https://www.halldrilling.com/) got a loan of $6.8 million.  They are "Drilling Oil and Gas Wells" service company.  They help people frack and extract oil and natural gas from the ground. This type of energy extraction has grown over the last decade, even as coal mining in the state has declined. They retained 387 jobs. Why might they have needed a loan? Oil and gas prices fell sharply in 2020 as people drove less and social distanced at home.  Those low prices made it harder for oil and gas companies to justify the cost of extracting it from the ground.  So it makes sense that a service provider like Hall would see less demand for its services. https://www.wvpublic.org/news/2020-03-19/coronavirus-and-slumping-prices-hit-ohio-valleys-oil-gas-sector

```{r}
wv_mining_ritchie <- ppp_wv_loans |>
  mutate(two_digit_naics = str_sub(naics_code,start=1L, end=2L)) |>
  filter(project_county_name == "RITCHIE") |>
  filter(two_digit_naics == "21") |>
  arrange(desc(amount)) |>
  head(1)

wv_mining_ritchie
```

-30-
