---
title: "lab_06"
author: "derek willis"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* [install Tabula](https://tabula.technology/)
* run existing code as directed (look for **Task**).
* write code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes in the lab_06 folder and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
```

## Get Our PDF

We'll be working with the [Maryland monthly voter registration update](https://elections.maryland.gov/pdf/vrar/2022_07.pdf) from the State Board of Elections. You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository).

**Task** Download the file from the above link and save it to your Downloads or labs folder.

## Setup Tabula

**Task** Start Tabula and load the PDF, then extract your tables

Start Tabula, then go to http://127.0.0.1:8080/ in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds.

This PDF has multiple possible tables to extract. We're going to make three dataframes: new registrations by party, removals by party, and current registrations by county. You will need to draw boxes around and export three CSV files (you could probably do this with just two files, but three is easier) that you will load into R. In Tabula, draw a box around the borders of those tables and click the "Preview & Export Extracted Data" button for each one. The export should look pretty clean. Let's export those CSVs (each one should be called `tabula-2022-07.csv` by default, so you will want to rename them to something better) to your lab_06/data folder.

## Cleaning up the data in R

From there, you will need to clean up the data, and in particular the headers. You can choose to include the headers from the PDF in your exported CSV files OR exclude them and add them when importing. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)):

**Task** Write code to read the CSV files into three dataframes called `party_registrations`, `party_removals` and `current_registrations`. You may have to skip 1 or more rows and you may want to supply easier-to-use column names that use underscores and don't begin with numbers. You also may need to ensure that columns containing numbers have a numeric datatype.

```{r}
# Write the code to load the CSV file here
```

Add a column that will contain the month for the update and populate it:

**Task** Write the code to add a `migration_type` column to the data and give it the value of "international"

```{r}
# Write the code to add the column here
```

## Answer questions

**Q1** 
You'll need to add and populate a column representing the percent of total using `mutate`.
**A1** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q2** Write a sentence or two that describes the data you produced in A1. Try to convey the most important idea.
**A2** [write your answer here]

**Q3** Which region & jurisdiction had the largest percentage change (either negative or positive) for international migration between July 2018 and July 2017? The formula for percentage change is easy to remember: (New-Old)/Old.
**A3** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q4** What's your best guess as to why these changes occurred, and in those places in particular?
**A4** [write your answer here]


## Back to Tabula

**Task** Extract second table from Tabula

Let's go to page 31 of the PDF, Table 2B, "Domestic Migration for Maryland's Jurisdictions, July 1, 2010 to July 1, 2019". In Tabula, hit the "Clear All Selections" button and then draw a box around that table's border and click the "Preview & Export Extracted Data" button. It should look pretty clean. Let's export that CSV to your lab_06/data folder. (and rename it to `tabula-md-statistical-handbook-domestic.csv`).

## Cleaning up the data in R

Let's load it into R, and in doing so we'll skip the first two rows and add our own headers that are cleaner:

**Task** Write code to read the CSV file into a dataframe called `domestic_migration`. As in the pre_lab, you may have to skip 1 or more rows and you may want to supply easier to use column names that use underscores. Your work on the international file should serve as a guide.

```{r}
# Write the code to load the CSV file here
```

Add a column for the type of migration ("domestic") and populate it:

**Task** Write the code to add a `migration_type` column to the data and populate it with the value of "domestic"

```{r}
# Write the code to add the column here
```

## Answer questions
**Q5** Which Maryland individual jurisdiction saw the largest net decrease in domestic migration overall?
**A5** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q6** How many regions & jurisdictions had net positive migration for July 2017, July 2018 and July 2019 (not just individually but all three together)?
**A6** [write your answer here]

```{r}
# Write the code to produce your answer here

```

**Q7** How would you describe this data? Is there a county or region that stands out, and why?
**A7** [write your answer here]
