---
title: "lab_01"
author: "Derek Willis"
date: "2024-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the slice of WinRed contribution data from West Virginia.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
wv_winred_contribs <- read_rds("lab_01.rds")
```

## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.

**Q1** How many rows and columns does this dataframe have?

**A1** 40,441 rows and 24 columns

```{r}

glimpse(wv_winred_contribs)

```

## Answer questions

**Q2** Which West Virginia city provided the largest number of contributions? How many?

**A2** Charleston, with 3,226.

```{r}
wv_winred_contribs |>
  group_by(city) |>
  summarise(
    total_contribs = n()
  ) |>
  arrange(desc(total_contribs))
```

**Q3** Which West Virginia cities accounted for the fewest contributions? How many of them, roughly?

**A3** Dozens of them, with 1 contribution.

```{r}
wv_winred_contribs |>
  group_by(city) |>
  summarise(
    total_contribs = n()
  ) |>
  arrange(total_contribs)
```

**Q4** What's your best educated guess to explain why the city identified in question 2 had more contributions compared to the cities identified in question 3? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.

**A4** Population, mostly, and Charleston is the capital. At the other end of the spectrum you have tiny towns spread across the state.

**Q5** What recipient got the largest total amount of money? How many contributions did they receive and what was the average donation?

**A5** Trump National Committee JFC, with \$463,944 from 14,063 contributions. The average was \$32.99.

```{r}
wv_winred_contribs |>
  group_by(memo_text) |>
  summarise(
    total_contribs = n(),
    total_amount = sum(amount),
    average_amount = mean(amount
    )
  ) |>
  arrange(desc(total_amount))
```

**Q6** What is that top recipient and what explains its popularity? Hint: this will require web research, you won't be able to answer this purely from the data set. You do not need to write any code for this question, so no codeblock is provided.

**A6** Trump National Committee JFC is the main fundraising organization for former President Donald Trump. It splits money between his campaign and other organizations. Trump won 68% of the vote in West Virginia in 2020.

**Q7**. Using the answer from A5 as a starting point, produce a summary table that has the following information for each recipient: \* count of contributions \* total amount of all contributions \* the average contribution amount \* the median contribution amount \* the smallest contribution amount \* the largest contribution amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings.

**A7**.

```{r}
wv_summary <- wv_winred_contribs |>
  group_by(memo_text) |>
   summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount),
    min_amount = min(amount),
    max_amount = max(amount)
  ) |>
  arrange(desc(total_amount))
```
