---
title: "lab_07"
author: "derek willis"
date: "2024-08-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key _only_ if you haven't already

```{r echo=FALSE}
census_api_key("YOUR API KEY", overwrite=TRUE)
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2022 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2022 5-year ACS and save them to a dataframe

```{r}
v20 <- load_variables(2020, "acs5", cache=TRUE)
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2020 5-year estimates? You will need to find the variable for median age first. 

**A1** Baltimore City, with an estimated median age of 36.0 years and a margin of error of 0.2 years.

```{r}
md_median_age <- get_acs(geography='county', variables=c(median_age='B01002_001'), state='MD', year=2022)
arrange(md_median_age, estimate)
```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. 

**A2** Talbot and Worcester counties

```{r}
md_median_age %>% filter(estimate >= 50)
```

## Ticket-Splitting

Ticket-splitting (voting for candidates from multiple parties) in today's political environment seems like a rarity, but it does happen, even in states like Maryland. For the remaining questions you'll be looking at cast vote records from Frederick County - Maryland's most competitive county - for the 2022 general election. You'll be trying to identify the zip codes where ticket-splitting is happening the most and to learn more about those places. To do that, you'll need to read in two CSV files, join them and then also bring in data from the Census API.

### Setup

Load the two CSV files in the `data` folder: one containing the ballot choices of Frederick voters in five races (governor, comptroller, attorney general and U.S. House) and the other containing information about Frederick's voting locations. You'll need to join them together.

```{r}

frederick_cvr <- read_csv("data/frederick_cvr.csv")
frederick_precincts <- read_csv("data/frederick_precincts.csv")

joined_frederick <- frederick_cvr |> left_join(frederick_precincts, join_by(precinct)) 
```

**Q3** Let's answer some basic questions about ticket-splitting: how many voters voted for different parties in the governor and comptroller's race? How many voted for the same party in both races? What percentage of all voters split their tickets in those races? In how many cases did voters choose the same party in all five contests? What percentage of all voters does that represent?

**A3** 11,871 Frederick voters, or 11 percent, split their tickets in the race. 89,916 voters, or nearly 84 percent, voted for the same way in all five races.

```{r}
joined_frederick |> filter(governor_lt_governor != comptroller) |> count()
joined_frederick |> filter(governor_lt_governor == comptroller) |> count()
pct <- 11871/107302 * 100

joined_frederick |> filter(governor_lt_governor == comptroller, governor_lt_governor == attorney_general, governor_lt_governor == u_s_senator, governor_lt_governor == representative_in_congress_6) |> count()

89916/107302 *100

```

**Q4** Now let's look for the zip codes where voters preferred Republicans in every race _except_ the governor's race, where Dan Cox lost to Wes Moore by a wide margin. Write code to generate a dataframe of zip codes with the number of those voters for each of those zip codes, ordering the dataframe so that the zip code with the most ticket-splitters appears first.

**A4** There are 23 zip codes in the county that had loyal Republicans who did not support Cox. 21702 had the most with 243.

```{r}
rep_no_cox_zips <- joined_frederick |> 
  filter(governor_lt_governor != 'REP', comptroller == 'REP', attorney_general == 'REP', u_s_senator == 'REP', representative_in_congress_6 == 'REP') |> 
  group_by(zipcode) |> 
  summarize(voters = n()) |> 
  arrange(desc(voters))
```


**Q5** Let's find out more about those zip codes from A4. Using tidycensus, get a dataframe of Maryland zip codes and their median ages, then join it to the dataframe you produced in A4. NOTE: you'll need to ensure that the two columns you use for the join are the same datatype. Summarize your findings; how would you describe the zip codes with more non-Cox Republicans compared to those with fewer? Where are those leading zip codes located?

**A5** The zip codes with the most non-Cox Republicans appear to have lower median ages than those with fewer such voters, although the difference isn't enormous. The top zip codes are centered in or just outside the city of Frederick.

```{r}
# get the zips with median income
md_zips <- get_acs(geography = "zcta",
              variables = ("B01002_001"),
              state = "MD",
              year = 2019)

# setup the join

rep_no_cox_zips <- rep_no_cox_zips |> 
  mutate(zipcode = as.character(zipcode))

# do the join

rep_no_cox_zips |> inner_join(md_zips, join_by(zipcode==GEOID))

```

