---
title: "lab_04"
author: "derek willis"
date: "8/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

library(tidyverse)
library(janitor)
```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use Maryland 2022 primary election results to do this, and our main question will be this: Which precincts had the most votes cast at them statewide, counting both the Republican and Democratic primaries? To answer this question, we'll need to first combine the two parties' primary precinct results into a single dataframe and then join it to data that has more information about each voting location.

First, load the two precinct results CSV files in the `data` folder, one each for the Democratic and Republican primaries, and then combine them into a single dataframe. You'll need to clean up the column names to make the data easier to work with - make sure you have loaded the library to do that.

```{r}
primary_dem <- read_csv("data/md_primary_precinct_dem.csv")
primary_rep <- read_csv("data/md_primary_precinct_rep.csv")

all_primary_precinct_results <- bind_rows(list(primary_dem, primary_rep))
all_primary_precinct_results <- all_primary_precinct_results %>% clean_names()
```

Next, let's read in the precinct voting location data and make sure you clean up the column names:

```{r}
precinct_voting_locations <- read_csv('data/precinct_voting_locations.csv') %>% clean_names()
```

Next join the two on the unique combination of precinct and county:

```{r}
# there's a problem - the `election_district_precinct` column in all_primary_precinct_results is numeric, so we have to convert it to character before we join

primary_precinct_results_with_voting_locations <- all_primary_precinct_results %>% left_join(precinct_voting_locations, by=c("election_district_precinct"="precinct_code", "county"="js_code"))
```

```{r}
all_primary_precinct_results <- all_primary_precinct_results %>% mutate(election_district_precinct = as.character(election_district_precinct))

# now we can join on precinct and county

primary_precinct_results_with_voting_locations <- all_primary_precinct_results %>% left_join(precinct_voting_locations, by=c("election_district_precinct"="precinct_code", "county"="js_code"))
```


## Answer questions

Q1. Look at the joined dataframe - are there any column names that are repeated, and what does that look like? What's the difference between the values in those repeated columns and what are your options for dealing with them? 

A1. There are two `county_name` columns, one from each data frame used in the join, so if R doesn't use those columns in the join itself (and we didn't because the content doesn't exactly match) it will include both and add `.x` and `.y` to their names to distinguish them. We could rename either or both of them and we could rename one and drop the other using select(-column_name).

Q2. Write code to produce a `votes` column which is the result of adding together `early_votes`, `election_night_votes`, `provisional_votes` and the mail-in votes, overwriting your dataframe so that it now has that `total_votes` column going forward. Then, separately write code to show each county and precinct and calculate the total votes in the Governor's race - bearing in mind that each precinct will have multiple rows in the dataframe, but you only want to see one row for each precinct. What is the name and voting location of the precinct that saw the most votes in the governor's race? 

A2. Precinct 15005 in Prince George's County, which votes at Wise High School.

```{r}
primary_precinct_results_with_voting_locations <- primary_precinct_results_with_voting_locations %>% 
  mutate(votes = early_votes + election_night_votes + provisional_votes + mail_in_ballot_1_votes + mail_in_ballot_2_votes)

primary_precinct_results_with_voting_locations %>% 
  filter(office_name == 'Governor / Lt. Governor') %>% 
  group_by(county_name.x, election_district_precinct, location_name) %>% 
  summarise(total_votes = sum(votes)) %>% 
  arrange(desc(total_votes))

```

Q3. Let's look at the average number of votes in each party's primary race for governor. Take the code you wrote for the last part of Q2, save the results to a new variable and then find the precincts with 0 total votes for governor. How many precincts are there, and which county has the most? What else would you need to know about these precincts to know whether this might be worth a story? Finally, write code to calculate the average and median number of total votes across all precincts (to do this, you will need to use the `ungroup()` function _before_ you do the math). Take a look at the the Annotated Code of Maryland's election laws (https://mgaleg.maryland.gov/2023RS/Statute_Web/gel/gel.pdf) - is there a limit on the number of voters assigned to a precinct? How might your answer impact wait times to vote?

A3. There are 12, and four of them are in Prince George's County. It would be useful to know how many registered voters are assigned to this precinct - if it's a sizeable amount (more than 25, say), that's probably worth looking into. The average number of votes across all precincts is 465, and the median is 418. Maryland has no limit on the number of voters assigned to a precinct (although many states do), so having 2,000+ voters at one location and 400 at another could mean that more voting machines and poll workers would be needed at the former location.

```{r}
gov_precinct_totals <- primary_precinct_results_with_voting_locations %>% 
  filter(office_name == 'Governor / Lt. Governor') %>% 
  group_by(county_name.x, election_district_precinct, location_name) %>% 
  summarise(total_votes = sum(votes)) %>% 
  arrange(desc(total_votes))

gov_precinct_totals %>% filter(total_votes == 0)

gov_precinct_totals %>% ungroup() %>% summarise(average = mean(total_votes), median = median(total_votes))

```

Q4. Choose your own adventure: I want you to come up with a question you could ask of one of the dataframes we've created for this lab. It could involve candidates, or counties or types of voting, but it cannot be a variation of a previous question (so, for example, you can't ask, "What's the average number of votes per precinct in the U.S. Senate primary?"). First write the question and then try to answer it by writing code. If you're not sure of the exact code, try googling (or ask me or Ramsay for advice) but, at a minimum, describe what you think you'd have to do to answer it and what the results would look like. That can help.
A4: 

```{r}

```

